---
title: "Census-to-CoC crosswalk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Census-to-CoC crosswalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
have_tc <- requireNamespace("tidycensus", quietly = TRUE)
```

```{r setup}
library(COCHomeless)
```

The crosswalk datasets link Census geography to HUD Continuum of Care (CoC)
areas, in both directions:

* `tract_coc2019` / `tract_coc2022` -- a hard assignment of each census tract to
  the CoC containing it.
* `tract_coc_wt2019` -- area weights for tracts that straddle a CoC border.
* `county_coc2019` -- county <-> CoC area weights (`w_coc` apportions a CoC down
  to counties; `w_county` aggregates a county up to CoCs).

The crosswalk procedure is due to Tom Byrne (Boston University); please credit
Byrne when using these data.

## National overview

```{r national}
nrow(tract_coc2019)                       # ~73k tracts assigned
length(unique(tract_coc2019$COCNUM))      # ~394 CoCs
# tracts that split across more than one CoC (area-weighted file)
sum(table(tract_coc_wt2019$GEOID) > 1)
```

## Focus: a California and a Washington CoC

List the counties that make up Seattle/King County (WA-500) and how much of the
CoC falls in each, using the area weights:

```{r wa-counties}
wa500 <- county_coc2019[county_coc2019$COCNUM == "WA-500", ]
wa500[order(-wa500$w_coc), c("fips", "w_coc")]
```

The tracts assigned to San Francisco's CoC (CA-501):

```{r ca-tracts}
ca501 <- tract_coc2019[tract_coc2019$COCNUM == "CA-501", ]
head(ca501[, c("GEOID", "fips", "COCNAME")])
nrow(ca501)
```

## Aggregating an ACS variable up to a CoC

Pull a tract-level American Community Survey variable and aggregate it to CoCs
using the area weights. (Requires a Census API key; not run here.)

```{r acs-rollup, eval = FALSE}
library(tidycensus)
library(dplyr)
# tract population for Washington
pop <- get_acs("tract", variables = "B01003_001", state = "WA",
               year = 2019, survey = "acs5")
tract_coc_wt2019 %>%
  inner_join(pop, by = c("GEOID")) %>%
  group_by(COCNUM, COCNAME) %>%
  summarise(coc_population = sum(estimate * w_tract), .groups = "drop") %>%
  arrange(desc(coc_population))
```

## CoC codes over time

CoCs merge and are renumbered. Roll a historical code forward to its surviving
CoC with `coc_mergers`:

```{r mergers}
head(coc_mergers)
resolve_coc <- function(code, mergers = coc_mergers) {
  repeat {
    hit <- mergers$coc_post[match(code, mergers$coc_pre)]
    if (is.na(hit) || hit == code) return(code)
    code <- hit
  }
}
resolve_coc("AR-502")
```
