---
title: "CoC Point-in-Time homeless counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CoC Point-in-Time homeless counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
have_map <- requireNamespace("sf", quietly = TRUE) &&
            requireNamespace("ggplot2", quietly = TRUE) &&
            requireNamespace("tigris", quietly = TRUE)
have_arima <- requireNamespace("imputeTS", quietly = TRUE) &&
              requireNamespace("forecast", quietly = TRUE) &&
              requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(COCHomeless)
```

The `hud2007`--`hud2024` data frames hold HUD's annual Point-in-Time (PIT)
estimate of the total ("Overall Homeless") population for each Continuum of Care
(CoC). The `coc2007`--`coc2024` objects hold the matching CoC boundaries as
`sf` polygons.

## National overview

National totals over the full series:

```{r national-trend}
years <- 2007:2025
totals <- sapply(years, function(y) sum(get(paste0("hud", y))$count))
plot(years, totals / 1e6, type = "b", pch = 19,
     xlab = "Year", ylab = "Total PIT homeless (millions)",
     main = "U.S. Point-in-Time homeless counts, 2007-2024")
```

The 2021 dip reflects HUD's pandemic waiver of the unsheltered count for many
CoCs, not a real decline.

## Sheltered, unsheltered and total, with ARIMA reconstruction of 2021

`pit_us` splits the national total into sheltered and unsheltered counts. The
2021 unsheltered count collapsed because of the waiver; the sheltered count held
up. We treat 2021 as missing and reconstruct it for each series with an ARIMA
model (Kalman smoothing via `imputeTS::na_kalman`), shown as hollow points.

```{r arima-2021, eval = have_arima, fig.height = 4.4}
library(ggplot2)
series <- c("total", "sheltered", "unsheltered")
yr <- pit_us$year
# ARIMA-impute the 2021 value of one series (2021 set to NA)
arima_2021 <- function(v) {
  x <- v; x[yr == 2021] <- NA
  as.numeric(imputeTS::na_kalman(ts(x, start = min(yr)), model = "auto.arima"))[yr == 2021]
}
obs <- do.call(rbind, lapply(series, function(s)
  data.frame(year = yr, series = s, value = pit_us[[s]])))
imp <- data.frame(year = 2021, series = series,
                  value = vapply(series, function(s) arima_2021(pit_us[[s]]), numeric(1)))

ggplot(obs, aes(year, value, color = series)) +
  geom_line() + geom_point(size = 1.3) +
  geom_point(data = imp, shape = 21, fill = "white", size = 3.2, stroke = 1.1) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "U.S. PIT homeless by shelter status, 2007-2025",
       subtitle = "Hollow points: ARIMA reconstruction of the waiver-affected 2021 count",
       y = NULL, x = NULL, color = NULL) +
  theme_minimal()
```

```{r arima-2021-table, eval = have_arima}
data.frame(series, observed_2021 = sapply(series, function(s) pit_us[[s]][yr == 2021]),
           arima_2021 = round(imp$value))
```

A national map of 2024, CoCs shaded by PIT count. Alaska and Hawaii are
repositioned with `tigris::shift_geometry()` for a compact lower-48 layout:

```{r national-map, eval = have_map, fig.height = 4.6}
library(sf); library(ggplot2); library(tigris)
# drop the far Pacific/Caribbean territories (shift_geometry only repositions
# AK/HI/PR) so the map stays centered on the lower 48 + insets
conus <- function(x) x[!x$ST %in% c("AS", "GU", "MP", "VI"), ]
m24 <- conus(merge(coc2024, hud2024, by.x = "COCNUM", by.y = "coc_num"))
m24s <- shift_geometry(m24)
ggplot(m24s) +
  geom_sf(aes(fill = count), color = NA) +
  scale_fill_viridis_c(trans = "log10", name = "PIT\nhomeless",
                       labels = scales::comma) +
  labs(title = "Point-in-Time homeless by CoC, 2024") +
  theme_void()
```

## A helper for the two state examples

```{r helper, eval = have_map}
# the 3 largest CoCs in a state in 2024, and a short label for the map
top_cocs <- function(st, n = 3) {
  s <- m24[m24$ST == st, ]
  s <- s[order(-s$count), ][seq_len(n), ]
  s$label <- paste0(s$COCNUM, "\n", substr(s$COCNAME, 1, 22))
  s
}
# state map: all CoCs shaded by count, top 3 labeled by name
state_map <- function(st, title) {
  s <- m24[m24$ST == st, ]
  tp <- top_cocs(st)
  ggplot(s) +
    geom_sf(aes(fill = count), color = "grey70") +
    geom_sf_text(data = tp, aes(label = label), size = 2.6, fontface = "bold") +
    scale_fill_viridis_c(trans = "log10", name = "PIT", labels = scales::comma) +
    labs(title = title) + theme_void()
}
# per-CoC time series 2007-2024 for a set of CoC numbers
coc_series <- function(coc_nums) {
  do.call(rbind, lapply(coc_nums, function(cc) data.frame(
    year = years, COCNUM = cc,
    count = sapply(years, function(y) {
      d <- get(paste0("hud", y)); v <- d$count[d$coc_num == cc]
      if (length(v)) v else NA_real_ }))))
}
```

## California

```{r ca-map, eval = have_map, fig.height = 5.5}
state_map("CA", "California: PIT homeless by CoC, 2024 (top 3 labeled)")
```

Year-by-year trend for California's three largest CoCs:

```{r ca-trend, eval = have_map, fig.height = 4.2}
ca_top <- top_cocs("CA")
ggplot(coc_series(ca_top$COCNUM), aes(year, count, color = COCNUM)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  scale_color_discrete(labels = setNames(substr(ca_top$COCNAME, 1, 28), ca_top$COCNUM)) +
  labs(title = "California: largest CoCs, 2007-2024", y = "PIT homeless",
       color = NULL) + theme_minimal()
```

## Washington

```{r wa-map, eval = have_map, fig.height = 5}
state_map("WA", "Washington: PIT homeless by CoC, 2024 (top 3 labeled)")
```

Year-by-year trend for Washington's three largest CoCs:

```{r wa-trend, eval = have_map, fig.height = 4.2}
wa_top <- top_cocs("WA")
ggplot(coc_series(wa_top$COCNUM), aes(year, count, color = COCNUM)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  scale_color_discrete(labels = setNames(substr(wa_top$COCNAME, 1, 28), wa_top$COCNUM)) +
  labs(title = "Washington: largest CoCs, 2007-2024", y = "PIT homeless",
       color = NULL) + theme_minimal()
```

CoCs merge and are renumbered over time; use `coc_mergers` to follow a CoC
through a merger (see `?coc_mergers`).
